---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

```{r message=FALSE}
###Load Libraries###
library(dplyr)
library(ggplot2)

##READ IN CSV###
x <- read.csv("activity.csv", stringsAsFactors = FALSE)

###Convert Date variable to Date vector; replace in original dataframe and ensure columns are named correctly###
date <- strptime(x$date, "%Y-%m-%d")
df <-data.frame(x[[1]], date, x[[3]])
colnames(df) <- c("steps", "date", "interval")

###Create new variable called dow, populate with name of day###
df <- mutate(df, dow = weekdays(date))
                                
###Convert names of days into factor with Weekend and Weekday; add new variable to moriginal data frame###                             
weekdayslist <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
daytype <- factor((df$dow %in% weekdayslist), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
df <- data.frame(df, daytype)



```



## What is mean total number of steps taken per day?

```{r}
###Use DPLYR package to create data frame grouped by date###
stepsbyday<- df %>%
        group_by(date) %>%
        summarize(dailysteps = sum(steps, na.rm=TRUE))

stepsbyday <- filter(stepsbyday, dailysteps > 0)

meandailysteps <- as.integer(mean(stepsbyday$dailysteps, na.rm=TRUE))
mediandailysteps <- as.integer(median(stepsbyday$dailysteps, na.rm=TRUE))

hist(stepsbyday$dailysteps, 
     main = "Histogram of Daily Steps", xlab="Total Steps by Day", ylab="Number of Days", 
     col="lightblue")

```

Mean daily steps = `r meandailysteps`  
Median daily steps = `r mediandailysteps`

## What is the average daily activity pattern?

``` {r}
intervalsteps <- df %>%
        group_by(interval) %>%
        summarize(stepsinterval = sum(steps, na.rm=TRUE),  meanintervalsteps = mean(steps, na.rm=TRUE) )

moststeps <- top_n(intervalsteps, 1, meanintervalsteps)
moststeps <- moststeps[[1]]


plot(intervalsteps$interval, intervalsteps$meanintervalsteps, type="l",
     main = "Average Steps Per Interval Period", xlab="Interval Period", ylab="Mean Steps during Interval", 
     col="red")
```

The 5-minute interval that contains the maximum number of steps, on average, across all the days in the dataset is: `r moststeps`


## Imputing missing values

```{r}

###Create new dataframe by replacing NA's with average number of steps of interval across all days### 
newdf <- mutate(df, steps = ifelse(is.na(steps) == "TRUE", intervalsteps$meanintervalsteps, steps))
                
                stepsbydayadj <- newdf %>%
                        group_by(date) %>%
                        summarize(dailysteps = sum(steps, na.rm=TRUE))

meandailystepsadj = as.integer(mean(stepsbydayadj$dailysteps, na.rm=TRUE))
mediandailystepsadj = as.integer(median(stepsbydayadj$dailysteps, na.rm=TRUE))

                
hist(stepsbydayadj$dailysteps, 
     main = "Histogram of Daily Steps", xlab="Total Steps by Day", ylab="Number of Days", 
     col="lightblue")
```

###Strategy for Imputing Data
1. Find all records with NA value in steps variable.
2. Replace that step value with the mean across all days for that interval.


Imputed Mean: `r meandailystepsadj`  
Mean w/o NA replacment: `r meandailysteps`  
Difference:  `r meandailystepsadj - meandailysteps`

Imputed Median: `r mediandailystepsadj`  
Median w/o NA replacment: `r mediandailysteps`  
Difference: `r mediandailystepsadj - mediandailysteps`

## Are there differences in activity patterns between weekdays and weekends?

```{r}

intervalstepsbydaytype <- df %>%
        group_by(daytype, interval) %>%
        summarize(meandailysteps = mean(steps, na.rm=TRUE)  )


qplot(interval, meandailysteps, data = intervalstepsbydaytype, facets = .~daytype) + 
        geom_line()+
        labs(list(title = "Steps by Interval Period -- Weekend vs Weekday", 
                  x = "Interval Period", y = "Total Steps"))

```
   
Yes, there are differences between weekdays and weekends.  
- There appears to be more consistent activity throughout the day on the weekend.  
- There appears to be a period of spiked activity in the morning on weekdays, perhaps a walk to work, followed by minimal activity until about 5-6pm, perhaps when the subject leaves work.  